The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2013-08-16 script/en/chapters/primes.tex
CT Script: Minor additions (Zhang and Helfgott).

% --------------------------------------------------------------------------
% \vskip +10 pt
\newpage
\subsection{Some unproven statements / conjectures / open questions about primes}
\begin{itemize}
  \item Goldbach\footnote{%
     Christian Goldbach\index{Goldbach, Christian}, German mathematician,
     Mar 18, 1690$-$Nov 20, 1764. }
     conjectured: Every even natural number greater than 2 can
     be represented as the sum of two prime numbers.

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{HT-GoldbachConjecture}{}
\subsection{The Goldbach conjecture}
\label{L-GoldbachConjecture}

@@

\hypertarget{HT-WeakGoldbachConjecture}{}
\paragraph*{The weak Goldbach conjecture\footnote{%
   It is also known as the odd or ternary Goldbach conjecture.
   } }%
\index{Goldbach conjecture!weak}%
\label{L-WeakGoldbachConjecture}%
\mbox{}

@@

\noindent This conjecture ist still (since more than 250 years) unproven.
A preliminary work for a prove could be the prove published recently\footnote{\url{http://arxiv.org/abs/1201.6656}, submitted on 31 Jan 2012 (v1), last revised 3 Jul 2012 (v4)} by Terence Tao\index{Tao, Terence} from the University of California. He proved that every \textbf{odd} natural number greater than 1 can be represented as the sum of \textbf{at most five} prime numbers.

% \vskip +55 pt   % Don't want the headline stay alone on the previous page.
\hypertarget{HT-StrongGoldbachConjecture}{}
\paragraph*{The strong Goldbach conjecture\footnote{%
   It is also known as the even or binary Goldbach conjecture.
   } }%
\index{Goldbach conjecture!strong}%
\label{L-StrongGoldbachConjecture}%
\mbox{}

@@

\noindent Computers have verified\footnote{%
     It is generally accepted today, that the\index{Goldbach conjecture}
     \textbf{Goldbach conjecture} is true, 
     i.~e.\ valid for all even natural numbers greater than $2$.
     In 1999, mathematician J\"org Richstein\index{Richstein 1999}
     from the computer sciences institute at the University of Giessen,
     studied even numbers up to 400 billion ($4*10^{14}$) and found no
     contradictory example (\cite{pr:Richstein1999}).\\
     In the meantime further progress was made: See\\
     \url{http://www.ieeta.pt/~tos/goldbach.html} by Tom·s Oliveira e Silva,\\
     \url{http://en.wikipedia.org/wiki/Goldbach's\_conjecture},\\
     \url{http://primes.utm.edu/glossary/page.php/GoldbachConjecture.html}.\\      
     Nevertheless, this does not provide us with general proof.\\
     The fact is that despite all efforts, Goldbach's conjecture has to 
     date not been proven. This leads one to believe that since the pioneer
     work of the Austrian mathematician Kurt G\"odel \index{G\""odel, Kurt}
     is well-known, not every true mathematical theorem is provable (see
     \url{http://www.mathematik.ch/mathematiker/goedel.html}).
     Perhaps Goldbach's conjecture was correct, but in any case the proof
     will never be found. Conversely, that will presumably also remain 
     unproven.}
     the Goldbach conjecture for all even numbers up to $3*10^{17}$
     (as at September 2012), but no general proof has yet been
     found.\footnote{%
     The English publisher {\em Faber} and the American publisher 
     {\em Bloomsbury} issued in 2000 the 1992 published book 
     ``Uncle Petros and Goldbach's Conjecture'' by Apostolos Doxiadis.
     It's the story of an old maths professor who fails to prove a more 
     than 250 year old puzzle. To boost the sales figures the English and
     American publishers have offered a prize of 1 million USD, if someone
     can prove the conjecture -- which should be published by 2004 in a 
     well-known mathematical journal.\\
     Surprisingly only British and American citizens are allowed to 
     participate.
     }${}^,$\footnote{%
     The theorem which has come closest so far to Goldbach's conjecture
     was proved by Chen Jing-Run in 1966 in a way which is somewhat hard
     to understand:
     Each even integer greater than $2$ is the sum of one prime and of 
     the product of two primes. E.g.: $20=5+3*5.$\\
     Most of the research about the Goldbach conjecture is collected in 
     the book: ``Goldbach Conjecture'', ed. Wang Yuan, 1984, World 
     scientific Series in Pure Maths, Vol. 4.
     }${}^,$\footnote{%
     Especially this conjecture makes it clear, that even today we do not
     have a complete understanding of the deeper connections between 
     addition and multiplication of natural numbers.}

As bigger an even number is, as more umso such binary Goldbach partitions can be found -- in average: For $4$ there is only one partition $2 + 2$; for $16$ there are two, $3 + 13$ and $5 + 11$. With $100$ there are six such partitions: $3 + 97$, $11 + 89$, $17 + 83$, $29 + 71$, $41 + 59$, $47 + 53$.\footnote{%
     Progress in research about the Goldbach conjectures is often a topic in mathematical press column, like \url{http://oumathclub.wordpress.com/}.
     }   % end_footnote

@@

% --------------------------------------------------------------------------
\vskip +35 pt
\subsection{Open questions about twin primes and prime cousins}

Twin primes are prime numbers whose difference is 2. 
Examples include 5 and 7, or 101 and 103,
or $1,693,965 \cdot 2^{66,443} \pm 1$.

@@

\noindent Open questions are:
\begin{itemize}
\item What is the number of twin primes: Are there infinitely many or only a
      limited number?\footnote{%
      Remark: {\bf Triplet} primes, however, only occur once: 3, 5, 7.  
      For all other sets of three consecutive uneven numbers, one of them is 
      always divisible by 3 and thus not a prime.
                              }
\item Does a formula exist for calculating the number of twin primes per interval?
\end{itemize}

@@

A big step towards the solution of this problem was made by Dan Goldston,
J\'{a}nos Pintz and Cem Yildirim in 2003.\footnote{%
    D. A. Goldston: ``Gaps Between Primes''\\
    \url{http://www.math.sjsu.edu/~goldston/OberwolfachAbstract.pdf}\\ 
    See also: 
    \begin{compactitem}
       \item D. A. Goldstone: "Are There Infinitely Many Twin Primes?"\\
            \url{http://www.math.sjsu.edu/~goldston/twinprimes.pdf}
       \item K. Soundararajan: "Small Gaps Between Prime Numbers:
            The Work Of Goldston-Pintz-Yildirim"\\
	    \url{http://www.ams.org/bull/2007-44-01/S0273-0979-06-01142-6/S0273-0979-06-01142-6.pdf}
    \end{compactitem}
    \vspace{-\baselineskip} % TODO: Geht das nicht anders?

                                               }
The three mathematicians were investigating the distribution of prime numbers.
They could proof, that \[ \liminf_{n\to\infty}{\frac{p_{n+1}-p_n}{\log{p_n}}}=0,\]
where $p_{n}$ denotes the $n$-th prime number. \\
This means that the smallest limit point ($\liminf$) of the sequence
$\frac{p_{n+1}-p_n}{\log{p_n}}$ equals zero.\\
A point is called limit point of a sequence, if there lie in any arbitrarily small
neighbourhood of that point infinitely many elements of the sequence.\\
$\log{p_n}$ is about the average distance between the prime $p_n$ and the next prime
$p_{n+1}$.\\
Hence, the term above implies, that there are infinitely many consecutive
primes with a gap between them which is arbitrarily small compared to the expected
average gap.

@@

xxxxxxxxxxxxxx

@@

\item Goldbach conjecture verification project von Tom·s Oliveira e Silva,
      \index{Goldbach project}\\ 
      \url{http://www.ieeta.pt/~tos/goldbach.html}

>>>>

% --------------------------------------------------------------------------
% \vskip +10 pt
\newpage
\subsection{Some unproven statements / conjectures / open questions about primes}
\begin{itemize}
  \item Goldbach\footnote{%
     Christian Goldbach\index{Goldbach, Christian}, German mathematician,
     Mar 18, 1690$-$Nov 20, 1764. }
     conjectured: Every even natural number greater than 2 can
     be represented as the sum of two prime numbers.\footnote{Please compare
     chapter \ref{L-GoldbachConjecture}.}

@@

% --------------------------------------------------------------------------
\vskip +50 pt %Zuerst nur +10, aber dann war der Titel "The strong Goldbach conjecture" allein unten auf der Seite.
\hypertarget{HT-GoldbachConjecture}{}
\subsection{The Goldbach conjecture}
\label{L-GoldbachConjecture}

@@

\hypertarget{HT-WeakGoldbachConjecture}{}
\paragraph*{The weak Goldbach conjecture\footnote{%
   It is also known as the \textbf{odd} or \textbf{ternary} Goldbach conjecture.
   } }%
\index{Goldbach conjecture!weak}%
\label{L-WeakGoldbachConjecture}%
\mbox{}

@@

\noindent This conjecture ist still (since more than 250 years) assumed
unproven.

\noindent Computers have verified the \textbf{weak Goldbach conjecture} for all odd
natural numbers up to: $4*10^{18}$ (simple check at April 2012) and
$4*10^{17}$ (double check at May 2013).\footnote{%
See \url{http://sweet.ua.pt/tos/goldbach.html} by Tom·s Oliveira e Silva
    }

Previous results proved, that the weak Goldbach conjecture is true for all
odd natural numbers larger $e^{3100}\approx 2 \times 10^{1346}$.

As with many famous conjectures in mathematics, there are also a number of
purported proofs of the Goldbach conjecture, none accepted by the mathematical
community yet.\footnote{%
One of them is in the paper from Shan-Guang Tan, submitted on 16 Oct 2011 (v1),
last revised 26 Jun 2013 (v11) which claims to even proof the strong Goldbach
conjecture.\\
v1 has the title ``A proof of the Goldbach conjecture''.\\
v11 has the title ``On the representation of even numbers as the sum and difference of two primes and the representation of odd numbers as the sum of an odd prime and an even semiprime and the distribution of primes in short intervals''.\\
See \url{http://arxiv.org/abs/1110.3465}.
   }

A preliminary work for a prove could be the prove published recently\footnote{\url{http://arxiv.org/abs/1201.6656}, submitted on 31 Jan 2012 (v1), last revised 3 Jul 2012 (v4)} by Terence Tao\index{Tao, Terence} from the University of California. He proved that every \textbf{odd} natural number greater than 1 can be represented as the sum of \textbf{at most five} prime numbers.

Considerable work has been done on Goldbach's weak conjecture, culminating in a 2013 claim by Harald Helfgott to fully prove the conjecture for all odd integers greater than 7.\footnote{%
Helfgott, H.A. (2013): ``Major arcs for Goldbach's theorem''.
\url{http://arxiv.org/abs/1305.2897}.\\
v2 of his 133 pages paper (14 Jun 2013) states, that the proof works starting at $10^{29}$ (instead of $10^{30})$. And for all numbers up to $10^{30}$ he claims, that he and David Platt proved at the computer that the weak Goldbach is valid.
\\
\url{http://truthiscool.com/prime-numbers-the-271-year-old-puzzle-resolved}.\\
\url{http://www.newscientist.com/article/dn23535-proof-that-an-infinite-number-of-primes-are-paired.html#.UgwhOpL0Ek0}.\\
\url{http://www.spiegel.de/wissenschaft/mensch/beweis-fuer-schwache-goldbachsche-vermutung-a-901111.html}.
}

% \vskip +55 pt   % Don't want the headline stay alone on the previous page.
\hypertarget{HT-StrongGoldbachConjecture}{}
\paragraph*{The strong Goldbach conjecture\footnote{%
   It is also known as the \textbf{even} or \textbf{binary} Goldbach conjecture.
   } }%
\index{Goldbach conjecture!strong}%
\label{L-StrongGoldbachConjecture}%
\mbox{}

@@

\noindent Computers have once verified\footnote{%
     It is generally accepted today, that the\index{Goldbach conjecture}
     \textbf{Goldbach conjecture} is true, 
     i.~e.\ valid for all even natural numbers greater than $2$.
     In 1999, mathematician J\"org Richstein\index{Richstein 1999}
     from the computer sciences institute at the University of Giessen,
     studied even numbers up to 400 billion ($4*10^{14}$) and found no
     contradictory example (\cite{pr:Richstein1999}).\\
     In the meantime further progress was made: See\\
     \url{http://sweet.ua.pt/tos/goldbach.html} by Tom·s Oliveira e Silva, \\
     \url{http://www.ieeta.pt/~tos/goldbach.html} by Tom·s Oliveira e Silva,\\
     \url{http://en.wikipedia.org/wiki/Goldbach's\_conjecture},\\
     \url{http://primes.utm.edu/glossary/page.php/GoldbachConjecture.html}.\\      
     Nevertheless, this does not provide us with general proof.\\
     The fact is that despite all efforts, Goldbach's conjecture has to 
     date not been proven. This leads one to believe that since the pioneer
     work of the Austrian mathematician Kurt G\"odel \index{G\""odel, Kurt}
     is well-known, not every true mathematical theorem is provable (see
     \url{http://www.mathematik.ch/mathematiker/goedel.html}).
     Perhaps Goldbach's conjecture was correct, but in any case the proof
     will never be found. Conversely, that will presumably also remain 
     unproven.}
     the Goldbach conjecture for all even numbers up to $4*10^{18}$
     (as at May 2013), but no general proof has yet been
     found.\footnote{%
     The English publisher {\em Faber} and the American publisher 
     {\em Bloomsbury} issued in 2000 the 1992 published book 
     ``Uncle Petros and Goldbach's Conjecture'' by Apostolos Doxiadis.
     It's the story of an old maths professor who fails to prove a more 
     than 250 year old puzzle. To boost the sales figures the English and
     American publishers have offered a prize of 1 million USD, if someone
     can prove the conjecture -- which should be published by 2004 in a 
     well-known mathematical journal.\\
     Surprisingly only British and American citizens are allowed to 
     participate.
     }${}^,$\footnote{%
     The theorem which has come closest so far to Goldbach's conjecture
     was proved by Chen Jing-Run in 1966 in a way which is somewhat hard
     to understand:
     Each even integer greater than $2$ is the sum of one prime and of 
     the product of two primes. E.g.: $20=5+3*5.$\\
     Most of the research about the Goldbach conjecture is collected in 
     the book: ``Goldbach Conjecture'', ed. Wang Yuan, 1984, World 
     scientific Series in Pure Maths, Vol. 4.
     }${}^,$\footnote{%
     Especially this conjecture makes it clear, that even today we do not
     have a complete understanding of the deeper connections between 
     addition and multiplication of natural numbers.}

As bigger an even number is, as more such binary Goldbach partitions can be found -- in average: For $4$ there is only one partition $2 + 2$; for $16$ there are two, $3 + 13$ and $5 + 11$. With $100$ there are six such partitions: $3 + 97$, $11 + 89$, $17 + 83$, $29 + 71$, $41 + 59$, $47 + 53$.\footnote{%
     Progress in research about the Goldbach conjectures is often a topic in mathematical press column, like \url{http://oumathclub.wordpress.com/}.
     }   % end_footnote

@@

% --------------------------------------------------------------------------
\vskip +35 pt
\subsection{Open questions about twin primes and cousin primes}

Twin primes\index{Twin prime} are prime numbers whose difference is exactly 2. 
Examples include 5 and 7, or 101 and 103,
or $1,693,965 \cdot 2^{66,443} \pm 1$.

@@

\noindent Open questions are:
\begin{itemize}
\item What is the number of twin primes: Are there infinitely many or only a
      limited number?\footnote{%
      Remark: {\bf Triplet} primes, however, only occur once: 3, 5, 7.  
      For all other sets of three consecutive odd numbers, one of them is 
      always divisible by 3 and thus not a prime.
                              }${}^,$\footnote{%
      The conjecture that there are infinite many twin primes is not obvious.
      It's known, that for large numbers in average the expected gap between
      primes is constantly growing and circa 2.3 times the number of decimal
      digits. For example, among 100-digit decimal numbers the expected gap
      between primes is in average 230.
      But this statement is true just on average -- often the gap is much
      bigger, often much smaller.
                               }${}^,$\footnote{%
      \url{http://en.wikipedia.org/wiki/Twin_Prime_Conjecture}
      }
\item Does a formula exist for calculating the number of twin primes per interval?
\end{itemize}

@@

A big step towards the solution of this problem was made by Dan Goldston,
J\'{a}nos Pintz and Cem Yildirim in 2003.\footnote{%
    D. A. Goldston: ``Gaps Between Primes''\\
    \url{http://www.math.sjsu.edu/~goldston/OberwolfachAbstract.pdf}\\ 
    See also: 
    \begin{compactitem}
       \item D. A. Goldstone: "Are There Infinitely Many Twin Primes?"\\
            \url{http://www.math.sjsu.edu/~goldston/twinprimes.pdf}
       \item K. Soundararajan: "Small Gaps Between Prime Numbers:
            The Work Of Goldston-Pintz-Yildirim"\\
	    \url{http://www.ams.org/bull/2007-44-01/S0273-0979-06-01142-6/S0273-0979-06-01142-6.pdf}
    \end{compactitem}
    \vspace{-\baselineskip} % TODO: Geht das nicht anders?
                                               }
The three mathematicians were investigating the distribution of prime numbers.
They could proof, that \[ \liminf_{n\to\infty}{\frac{p_{n+1}-p_n}{\log{p_n}}}=0,\]
where $p_{n}$ denotes the $n$-th prime number. \\
This means that the smallest limit point ($\liminf$) of the sequence
$\frac{p_{n+1}-p_n}{\log{p_n}}$ equals zero.\\
A point is called limit point of a sequence, if there lie in any arbitraril
small neighbourhood of that point infinitely many elements of the sequence.\\
$\log{p_n}$ is about the average distance between the prime $p_n$ and the next
prime $p_{n+1}$.\\
Hence, the term above implies, that there are infinitely many consecutive
primes with a gap between them which is arbitrarily small compared to the
expected average gap.

@@

In May 2013 the results of Yitang Zhang\index{Zhang, Yitang} became
known.\footnote{%
    Erica Klarreich (May 19, 2013): ``Unheralded Mathematician Bridges the
    Prime Gap''\\
    \url{https://www.simonsfoundation.org/quanta/20130519-unheralded-mathematician-bridges-the-prime-gap/}}
Zhang proved, that there are infinitely many ``cousin primes'', or more
concretely that there is some number $N$ smaller than 70 million such that there
are inifinitely many pairs of primes that differ by N.\footnote{%
Whereas the gap between the primes of a twin prime is exactly $2$, do cousin
primes\index{Cousin prime} denote two primes, which have a gap between them,
which has a value of a bigger, even, but finite number.
      }${}^,$\footnote{%
This is close to the conjecture stated in 1849 by the French mathematician
Alphonse de Polignac that there are infinitely many prime pairs for any
possible even finite gap, not just $2$.
       }

@@

\item Goldbach conjecture verification project by Tom·s Oliveira e Silva,
      \index{Goldbach project}\\ 
      \url{http://sweet.ua.pt/tos/goldbach.html}
